### 2017-10-02 chapters/introduction.tex
Incorporate Sams suggestions

\section{Aim of thesis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% THE PAST %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Cardiovascular diseases are the leading cause of death in the
world, responsible for more than 17.3 million deaths annually, a
number that is expected to reach 23.6 million by 2030
\cite{writing2016heart}. Advancements in detection and treatment
have reduced morbidity and mortality, but knowledge of the underlying
mechanisms for the development of the disease is still very poor.
While some patients respond very well to a therapy, others suffering
from the same symptoms might not respond at all.
Our understanding of these mechanisms has been improved through
development of advance imaging techniques and through
computational modeling. While imaging play a central role in
assessment of cardiac structure and kinematics, biophysical models can
help us to understand causality in this multiscale environment.

@@

A paradigm shift in the field of cardiac mechanics modeling came in
the late 1980s at the University of Auckland. Here the first models
based on a continuum description of the heart were developed
\cite{hunter1988analysis}, which is still in use today.

@@

Today, advanced imaging techniques allow us to acquire detailed
mechanical information about the heart of the individual. In addition,
advanced numerical algorithms combined with high performance computing
power enable us to get more and more detailed simulation of a patients
heart.

@@

The need for building adequate patient-specific models that captures
the geometrical information observed in the images as well as the
underlying biophysical processes, is recognized as one of the key
challenges in modern bioengineering \cite{hunter2010vision}. This is
also reflected in the exponential increase in the yearly number of
publications on the topic \cite{sack2016personalised}.
In the future, models could be used to as a diagnostic tool to extract
features from a patients heart that are otherwise impossible to extract from
images. Eventually, models could potentially predict the outcome of
different treatment strategies and guide clinicians in the decision making.

@@

Another question is how to fuse data observed in medical images with
computational models, in order to make models specific to the
individual. These questions has to be addressed before we can take the
next step into personalized simulations for \emph{in silico} medicine,
and use computer simulations to guide treatment, and predict outcome.

@@

Some of these unanswered question serve as motivation for the work
conducted in this thesis. We focus purely on the mechanical modeling
of the heart and the aim has been to develop a framework for
building mechanical models that are personalized at the level of the
individual.

The mechanics of the heart can be described via equations of nonlinear
elasticity which is traditionally formulated using a continuum
mechanics approach. This approach has been extensively used in other
parts of engineering, and numerical solutions to these partial
differential equations (PDE) are solved using the finite element
method (FEM) with the open source software package FEniCS \cite{logg2012automated}.

These mechanical models contain model parameters which may vary from
individual to individual, and patient specificity can be obtained by
fitting these model parameters to additional clinical data. This
process is often referred to as data assimilation, a field that has
its root in meteorology where one want to make a weather forecast based
on observations from strategically placed weather stations. The idea
is to construct a functional which represent the mismatch between the
model and the observations, and by changing the model parameters we
also change the value of this mismatch functional. Now, what we want to do is to
minimize this functional using these model parameters as control. At
the same time, the governing PDE, which depend upon these model
parameters, has to remain true. We therefore often
refer to such problems as PDE-constrained optimization problems.  
Solving such problems is challenging because evaluation of the
mismatch functional requires the solution of a nonlinear PDE, which is
computationally expensive, and we therefore want to keep the number of
functional evaluations at a minimum. A good way of doing this is
traverse the parameter space along the gradient descent, which is
guaranteed to bring us towards a local minimum. However, this requires
calculation of the functional gradient, or the derivative of the
mismatch functional with respect to the model parameters. As we will
see, the functional gradient can be computed efficiently by
considering the adjoint equation, and recent development in numerical
software tools \cite{farrell2013automated} allow us to exploit this
also numerically.

@@

The remaining introduction in this thesis is organized as follows. In
Section \ref{sec:intro_physiology} we briefly explain some concepts in
cardiac anatomy and physiology which is used throughout the
thesis. This section ends with a description of the different cardiac
diseases encountered in this thesis. In Section
\ref{sec:intro_mechanical} we give an introduction to continuum
mechanics and mechanical modeling of the heart. The basic equations of
elasticity as well as the constitutive relations for the cardiac
tissue, and some numerical considerations are covered. In Section 
\ref{sec:intro_personalization} we explain how to personalize a
mechanical model using clinical data. We start by explaining the
different imaging modalities used in this thesis, and how finite
element meshes are created from images. We also explain how to find a
more realistic reference geometry than the one obtained from medical
images. At the end we introduce data assimilation, and how to compute
the functional gradient using the adjoint method. Finally in Section
\ref{sec:summary} we summarize the research papers that make up this
thesis, and make some closing remarks and suggestions for future
directions.

>>>>

\section{Aim of thesis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% THE PAST %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Cardiovascular diseases are the leading cause of death in the
world, responsible for more than 17.3 million deaths annually, a
number that is expected to reach 23.6 million by 2030
\cite{writing2016heart}. Advancements in detection and treatment
have reduced morbidity and mortality, but knowledge of the underlying
mechanisms for the development of the disease, is still very poor.
Consequently, while some patients respond very well to a therapy, others suffering
from the same symptoms might not respond at all.
Our understanding of these mechanisms has been improved through
development of advance imaging techniques and through
computational modeling. While imaging plays a central role in
assessment of cardiac structure and kinematics, biophysical models can
help us to understand causality in this multiscale environment.

@@

A paradigm shift in the field of cardiac mechanics modeling came in
the late 1980s at the University of Auckland. Here the first models
based on a continuum description of the heart were developed
\cite{hunter1988analysis}, which is still are use today.

@@

Today, advanced imaging techniques allow us to acquire detailed
mechanical information about the heart of an individual. In addition,
advanced numerical algorithms combined with high performance computing
power enable us to get more and more detailed simulation of a patients
heart.

@@

The need for building adequate patient-specific models that captures
the geometrical information observed in the images, as well as the
underlying biophysical processes, is recognized as one of the key
challenges in modern bioengineering \cite{hunter2010vision}. This is
also reflected in the exponential increase in the yearly number of
publications on the topic \cite{sack2016personalised}.
In the future, models could be used to as a diagnostic tool to extract
features of a patient's heart that are otherwise impossible to extract from
images. Eventually, models could potentially predict the outcome of
different treatment strategies and guide clinicians in the decision making.

@@

Another question is how to fuse data observed in medical images with
computational models, in order to make models specific to the
individual. These questions have to be addressed before we can take the
next step into personalized simulations for \emph{in silico} medicine,
and use computer simulations to guide treatment, and predict outcome.

@@

Some of these unanswered questions serve as motivation for the work
conducted in this thesis. We focus purely on the mechanical modeling
of the heart and the aim has been to develop a framework for
building mechanical models that are personalized at the level of the
individual.

The mechanics of the heart can be described via equations of nonlinear
elasticity which are traditionally formulated using a continuum
mechanics approach. This approach has been extensively used in other
parts of engineering, and in this work the numerical solutions to these partial
differential equations (PDE) are solved using the finite element
method (FEM) with the open source software package FEniCS \cite{logg2012automated}.

These mechanical models contain model parameters which may vary from
individual to individual, and patient specificity can be obtained by
fitting these model parameters to additional clinical data. This
process is often referred to as data assimilation, a field that has
its root in meteorology where one wants to make a weather forecast based
on observations from strategically placed weather stations. The idea
is to construct a functional which represent the mismatch between the
model and the observations, and by changing the model parameters we
also change the value of this mismatch functional. Now, what we want to do is to
minimize this functional using these model parameters as controls. At
the same time, the governing PDE, which depend upon these model
parameters, has to remain true. We therefore often
refer to such problems as PDE-constrained optimization problems.  
Solving such problems is challenging because evaluation of the
mismatch functional requires the solution of a nonlinear PDE, which is
computationally expensive, and we therefore want to keep the number of
functional evaluations at a minimum. A good way of doing this is
traverse the parameter space along the gradient descent, which is
guaranteed to bring us towards a local minimum. However, this requires
calculation of the functional gradient, or the derivative of the
mismatch functional with respect to the model parameters. As we will
see, the functional gradient can be computed efficiently by
considering the adjoint equation, and recent development in numerical
software tools \cite{farrell2013automated} allow us to exploit this
also numerically.

@@

The remaining introduction in this thesis is organized as follows. In
Section \ref{sec:intro_physiology} we briefly explain some concepts in
cardiac anatomy and physiology which are used throughout the
thesis. This section ends with a description of the different cardiac
diseases encountered in this thesis. In Section
\ref{sec:intro_mechanical} we give an introduction to continuum
mechanics and mechanical modeling of the heart. The basic equations of
elasticity as well as the constitutive relations for the cardiac
tissue, and some numerical considerations are covered. In Section 
\ref{sec:intro_personalization} we explain how to personalize a
mechanical model using clinical data. We start by explaining the
different imaging modalities used in this thesis, and how finite
element meshes are created from images. We also explain how to find a
more realistic reference geometry than the one obtained from medical
images. At the end we introduce data assimilation, and how to compute
the functional gradient using the adjoint method. Finally in Section
\ref{sec:summary} we summarize the research papers that make up this
thesis, and make some closing remarks and suggestions for future
directions.
